Generalized spaces for constructive algebra

### 2020-04-06 paper-herrsching2019/main.tex
McCoy's theorem

\newcommand{\A}{\mathcal{A}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\C}{\mathcal{C}}
\newcommand{\M}{\mathcal{M}}
\renewcommand{\AA}{\mathbb{A}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\G}{\mathcal{G}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\GG}{\mathbb{G}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\K}{\mathcal{K}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\TT}{\mathbb{T}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\CC}{\mathbb{C}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\bbb}{\mathfrak{b}}
\newcommand{\ppp}{\mathfrak{p}}
\newcommand{\qqq}{\mathfrak{q}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\fff}{\mathfrak{f}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\defeqv}{\vcentcolon\equiv}
\newcommand{\Sh}{\mathrm{Sh}}
\newcommand{\GL}{\mathrm{GL}}
\newcommand{\ab}{\mathrm{ab}}
\newcommand{\Zar}{\mathrm{Zar}}
\newcommand{\op}{\mathrm{op}}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Eff}{\mathrm{Ef{}f}}
\newcommand{\Sch}{\mathrm{Sch}}
\newcommand{\Aff}{\mathrm{Aff}}
\newcommand{\Ring}{\mathrm{Ring}}
\newcommand{\LocRing}{\mathrm{LocRing}}
\newcommand{\LRS}{\mathrm{LRS}}
\newcommand{\Hom}{\mathrm{Hom}}
\newcommand{\Spec}{\mathrm{Spec}}
\newcommand{\Max}{\mathrm{Max}}
\newcommand{\Gal}{\mathrm{Gal}}
\newcommand{\Rad}{\mathrm{Rad}}
\newcommand{\Idl}{\mathrm{Idl}}
\newcommand{\lra}{\longrightarrow}
\newcommand{\RelSpec}{\operatorname{Spec}}
\renewcommand{\_}{\mathpunct{.}}
\newcommand{\?}{\,{:}\,}
\newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
\newcommand{\ul}[1]{\underline{#1}}
\newcommand{\affl}{\ensuremath{{\ul{\ensuremath{\AA}}^1}}}
\newcommand{\Ll}{\text{iff}}
\newcommand{\inv}{inv.\@}
\newcommand{\seq}[1]{\mathrel{\vdash\!\!\!_{#1}}}
\newcommand{\hg}{\mathbin{:}}  % homogeneous coordinates
\newcommand{\brak}[1]{{\llbracket{#1}\rrbracket}}
\newcommand{\pt}{\mathrm{pt}}
\newcommand{\Pt}{\mathrm{Pt}}
\newcommand{\Loc}{\mathrm{Loc}}
\newcommand{\Top}{\mathrm{Top}}
\newcommand{\BPIT}{\textsc{bpit}\xspace}
\newcommand{\notnot}{\emph{not not}\xspace}
\newcommand{\negg}{\neg\!\!\!\neg}
\newcommand{\bott}{\bot\!\!\!\!\bot}

@@

%\begin{proposition}[McCoy's theorem]
%Let~$M \in A^{n \times m}$ be an injective matrix over a reduced ring~$A$.
%Then~$(\Lambda^m M)$, the ideal of~$m$-minors of~$M$, is regular.
%\end{proposition}
%
%\begin{proof}Let~$x \in A$ such that~$x \cdot (\Lambda^M) = (0)$ be given. We
%will show~$x = 0$~in~$A$ by verifying~$x = 0$ in~$A^\sim$ under the sheaf
%semantics.
%
%To this end, it is prudent to introduce a variant of the usual
%negation~$\neg\varphi \equiv (\varphi \Rightarrow \bot)$, namely~$\negg\varphi
%\equiv (\varphi \Rightarrow a = 0)$. Our claim is then~$\bott \defeq
%\negg\top$. As usual, the double negation of the law of excluded middle is
%available, that is we have
%\[ \forall x\?A^\sim\_ \negg\negg\bigl((\exists y\?A\_ xy = 1) \vee \negg(\exists
%y\?A\_ xy = 1)\bigr), \]
%and as our goal is~$\bott$, we may use with law of excluded middle a finite
%number of times.
%
%Repeatedly exploiting that any matrix element is invertible or
%$\negg\text{invertible}$ and applying elementary row and column
%transformations, we may assume that~$M$ is of the
%form~$\left(\begin{smallmatrix}I&0\\0&N\end{smallmatrix}\right)$ with all
%elements of~$N$ $\negg\text{invertible}$.
%
%By Remark~\ref{XXX},
%\end{proof}
% XXX

@@

@@

\begin{exercise}{A generalization of the field property}%
Let~$A$ be a ring. We extend the sheaf semantics to infinitary first-order
formulas by adding the following clauses to
Table~\ref{table:algebraic-kripke-joyal}:
\begin{center}\tablefont
\begin{tabular}{@{}l@{\ \ }c@{\ \ }l@{}}
  $f \models \bigwedge_{k \in K} \varphi_k$ &iff&
    for all indices~$k \in K$, $f \models \varphi_k$ \\
  $f \models \bigvee_{k \in K} \varphi_k$ &iff&
    there exists a partition~$f^n = fg_1 + \cdots + fg_m$ such that, \\
  &&\quad for each~$i$, there is an index~$k \in K$ such that~$fg_i \models \varphi_k$
\end{tabular}
\end{center}
Theorem~\ref{thm:basic-properties-sheaf-semantics-algebraic}
also applies to this extension, and moreover, this extension in sound with
respect to infinitary intuitionistic logic.
\begin{alphlist}[(c)]
\item Prove that:
\[ 1 \models \forall f\?A^\sim\_ \forall x\?A^\sim\_
  \bigl((\exists z\?A^\sim\_ fz = 1) \Rightarrow x = 0\bigr) \Longrightarrow \bigvee_{n \in \NN} (f^n x = 0). \]
\item Deduce that~$A^\sim$ is ``almost a field'' in the sense that any nonunit
is nilpotent.
\item Show that~$A$ is reduced iff~$A^\sim$ is.
\end{alphlist}
\end{exercise}

>>>>

\newcommand{\A}{\mathcal{A}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\C}{\mathcal{C}}
\newcommand{\M}{\mathcal{M}}
\renewcommand{\AA}{\mathbb{A}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\G}{\mathcal{G}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\GG}{\mathbb{G}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\K}{\mathcal{K}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\TT}{\mathbb{T}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\CC}{\mathbb{C}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\bbb}{\mathfrak{b}}
\newcommand{\ppp}{\mathfrak{p}}
\newcommand{\qqq}{\mathfrak{q}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\fff}{\mathfrak{f}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\defeqv}{\vcentcolon\equiv}
\newcommand{\Sh}{\mathrm{Sh}}
\newcommand{\GL}{\mathrm{GL}}
\newcommand{\ab}{\mathrm{ab}}
\newcommand{\Zar}{\mathrm{Zar}}
\newcommand{\op}{\mathrm{op}}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Eff}{\mathrm{Ef{}f}}
\newcommand{\Sch}{\mathrm{Sch}}
\newcommand{\Aff}{\mathrm{Aff}}
\newcommand{\Ring}{\mathrm{Ring}}
\newcommand{\LocRing}{\mathrm{LocRing}}
\newcommand{\LRS}{\mathrm{LRS}}
\newcommand{\Hom}{\mathrm{Hom}}
\newcommand{\Spec}{\mathrm{Spec}}
\newcommand{\Max}{\mathrm{Max}}
\newcommand{\Gal}{\mathrm{Gal}}
\newcommand{\Rad}{\mathrm{Rad}}
\newcommand{\Idl}{\mathrm{Idl}}
\newcommand{\lra}{\longrightarrow}
\newcommand{\RelSpec}{\operatorname{Spec}}
\renewcommand{\_}{\mathpunct{.}}
\newcommand{\?}{\,{:}\,}
\newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
\newcommand{\ul}[1]{\underline{#1}}
\newcommand{\affl}{\ensuremath{{\ul{\ensuremath{\AA}}^1}}}
\newcommand{\Ll}{\text{iff}}
\newcommand{\inv}{inv.\@}
\newcommand{\seq}[1]{\mathrel{\vdash\!\!\!_{#1}}}
\newcommand{\hg}{\mathbin{:}}  % homogeneous coordinates
\newcommand{\brak}[1]{{\llbracket{#1}\rrbracket}}
\newcommand{\pt}{\mathrm{pt}}
\newcommand{\Pt}{\mathrm{Pt}}
\newcommand{\Loc}{\mathrm{Loc}}
\newcommand{\Top}{\mathrm{Top}}
\newcommand{\ann}{\operatorname{ann}}
\newcommand{\BPIT}{\textsc{bpit}\xspace}
\newcommand{\notnot}{\emph{not not}\xspace}
\newcommand{\negg}{\neg\!\!\!\neg}
\newcommand{\bott}{\bot\!\!\!\!\bot}

@@

@@

\paragraph{McCoy's theorem.}\label{par:mccoy}
It is a familiar fact from linear algebra that an injective~$(n \times
m)$-matrix~$M$ over a geometric field has at least one invertible~$m$-minor.
Like many proofs in linear algebra, the proof proceeds by first applying
elementary row and column transformations to put~$M$ into the form~$N \defeq
\left(\begin{smallmatrix}I&0\\0&M'\end{smallmatrix}\right)$. This step exploits
that any element is zero or invertible. All entries in the submatrix~$M'$ are
zero, and since~$M$ is injective, the number of columns of~$M'$ is zero.
Hence~$N = \left(\begin{smallmatrix}I\\0\end{smallmatrix}\right)$ has an
invertible~$m$-minor, and thus~$M$ as well.

A version of \emph{McCoy's theorem} states that this fact
generalizes to arbitrary rings:

\begin{quote}
\textbf{Theorem.} Let~$M \in A^{n \times m}$ be an injective matrix.
Then~$(\Lambda^m M)$, the ideal of~$m$-minors of~$M$, is regular.
\end{quote}

The sheaf-theoretic reduction technique allows us to give a constructive proof
of McCoy's theorem which is close to the linear algebra proof. In case that the
ring~$A$ is reduced, we can even copy the original proof almost word for word;
for the general case, we require the following lemma.

\begin{quote}
\textbf{Lemma.} Let~$M \in A^{n \times m}$ be an injective matrix with~$m > 0$. Assume that all entries
of~$M$ are nilpotent. Then~$1 = 0$ in~$A$.

\textbf{Proof of the lemma.} Let~$(M) \subseteq A$ be the ideal generated by the entries
of~$M$. The assumption implies that there is a number~$K \in \NN$ such that~$(M)^K = (0)$.
If~$K \geq 1$, then also~$(M)^{K-1} = (0)$, since for~$x \in (M)^{K-1}$ the
coefficients of~$M (x,0,\ldots,0)^t$ are elements of~$(M)^K$, hence zero,
and~$M$ is injective. We may thus assume~$K = 0$. Hence~$1 \in (M)^0 = (0)$. \qed
\end{quote}
The proof now proceeds as follows.
\begin{quote}
\textbf{Proof of the theorem.} Let~$x \in A$ be such that~$x (\Lambda^m M) = (0)$. We are to
show~$x = 0$. By passing to the factor ring~$A/\ann(x)$, we may
assume~$(\Lambda^m M) = (0)$ and~$x = 1$; the new goal is to verify~$1 = 0$. We
do this by validating~$\bot$ in the sheaf semantics.

We may employ the law of excluded middle a finite number of times because we
are to verify~$\bot$. By repeatedly exploiting that any matrix entry is
invertible or not, we may use row and column transformations to put~$M$ into
the form~$N \defeq \left(\begin{smallmatrix}I&0\\0&M'\end{smallmatrix}\right)$ where all
entries of~$M'$ are not invertible. By~XXX, the entries of~$M'$ are
nilpotent, and since~$M'$ is injective and~$1 \neq 0$, the preceding lemma
shows that the number of columns of~$M'$ is zero.

Hence~$N = \left(\begin{smallmatrix}I\\0\end{smallmatrix}\right)$ and~$(1) =
(\Lambda^m N) = (\Lambda^m M) = (0)$, a contradiction. \qed
\end{quote}

\begin{remark}Not of pragmatic, but of logical interest is that the proof can
also be recast as follows. Instead of reducing to the factor ring~$A/\ann(x)$,
we continue considering~$A$. Working under the sheaf sematics, we then
introduce a variant of the usual negation~$\neg\varphi \equiv (\varphi
\Rightarrow \bot)$, namely~$\negg\varphi \equiv (\varphi \Rightarrow a = 0)$.
We may still use the law of excluded middle for this new negation and hence
exploit that any element~$x$ of~$A^\sim$ is invertible or ``not'' invertible. By
Exercise~\ref{ex:gen-field-property}, this amounts to saying that some power
of~$x$ annihilates~$a$. The remainder of the proof carries over with minor
modifications.
\end{remark}

As with the other examples, the proof can mechanically be unwound to obtain a proof
which is fully explicit:

\begin{quote}
\textbf{Proof.} Let~$x \in A$ be such that~$x (\Lambda^m M) = (0)$. We are to
show~$x = 0$. By passing to the factor ring~$A/\ann(x)$, we may
assume~$(\Lambda^m M) = (0)$ and~$x = 1$; the new goal is to verify~$1 = 0$.

If~$m = 0$, the claim is immediate because in this case~$(1) = (\Lambda^m M) =
(0)$. So assume~$m > 0$.

Over the localized ring~$A[M_{ij}^{-1}]$, the matrix entry~$M_{ij}$ is
invertible and hence~$M$ can be put into the
form~$\left(\begin{smallmatrix}1&0\\0&M'\end{smallmatrix}\right)$ by elementary
row and column transformation. The matrix~$M'$ is still injective,
and~$(\Lambda^{m-1} M') = (\Lambda^m M)$, hence we may assume by induction
that~$1 = 0$ in~$A[M_{ij}^{-1}]$.

Hence all entries of~$M$ are nilpotent and the claim follows from the preceding
lemma. \qed
\end{quote}

Other short and explicit proofs of McCoy's theorem are given in
Refs.~\refcite{blocki:mccoy},
\cite[Theorem~5.22]{lombardi-quitte:constructive-algebra} and
\cite[Exercise~5.23A(3)]{lam:exercises}. The proof in
Ref.~\cite[Proposition~III.8.2.3]{bourbaki:alg} employs the law of excluded
middle. We were not able to discern that these proofs are just different
coatings of the same idea that ours is based on; these proofs all employ
computations with Cramer-like formulas, while our proof stresses the similarity
to the field case.

It is on purpose that the (sheaf-theoretic version of) our proof contains case
distinctions, to closely mimick the proof in the field case. It is also
possible to do without, as already our unrolled proof demonstrates. Thierry
Coquand and Claude Quitté gave a beautiful proof without case distinctions an
application of a local--global
principle~\cite[Theorem~2.4]{coquand-quitte:constructive-finite-free-resolutions};
if desired, their proof can also be cast in the sheaf-theoretic framework
(Exercise~\ref{ex:local-global}).

@@

\begin{exercise}{A generalization of the field property}%
\label{ex:gen-field-property}%
Let~$A$ be a ring. We extend the sheaf semantics to infinitary first-order
formulas by adding the following clauses to
Table~\ref{table:algebraic-kripke-joyal}:
\begin{center}\tablefont
\begin{tabular}{@{}l@{\ \ }c@{\ \ }l@{}}
  $f \models \bigwedge_{k \in K} \varphi_k$ &iff&
    for all indices~$k \in K$, $f \models \varphi_k$ \\
  $f \models \bigvee_{k \in K} \varphi_k$ &iff&
    there exists a partition~$f^n = fg_1 + \cdots + fg_m$ such that, \\
  &&\quad for each~$i$, there is an index~$k \in K$ such that~$fg_i \models \varphi_k$
\end{tabular}
\end{center}
Theorem~\ref{thm:basic-properties-sheaf-semantics-algebraic}
also applies to this extension, and moreover, this extension in sound with
respect to infinitary intuitionistic logic.
\begin{alphlist}[(c)]
\item Prove that:
\[ 1 \models \forall f\?A^\sim\_ \forall x\?A^\sim\_
  \bigl((\exists z\?A^\sim\_ fz = 1) \Rightarrow x = 0\bigr) \Longrightarrow \bigvee_{n \in \NN} (f^n x = 0). \]
\item Deduce that~$A^\sim$ is ``almost a field'' in the sense that any nonunit
is nilpotent.
\item Show that~$A$ is reduced iff~$A^\sim$ is.
\end{alphlist}
\end{exercise}

\begin{exercise}{A local--global principle}%
\label{ex:local-global}%
Let~$A$ be a ring. We introduce an operator~$\nabla$ by
\[ \nabla\varphi \defeqv \bigl(\forall s\?A^\sim\_ ((\varphi \Rightarrow s = 0)
\Rightarrow s = 0\bigr).\]
\begin{alphlist}[(f)]
\item Verify that~$\nabla$ is a \emph{local operator}, that is verify that
\[ (\varphi \Rightarrow \nabla\varphi), \quad
(\nabla\nabla\varphi \Rightarrow \nabla\varphi) \quad\text{and}\quad
((\nabla\varphi \wedge \nabla\psi) \Leftrightarrow \nabla(\varphi \wedge \psi))
\]
hold under the sheaf semantics.
\item Show that equality is~$\nabla$-stable in that
\[ 1 \models \forall x\?A^\sim\_ \forall y\?A^\sim\_ \nabla(x = y) \Rightarrow x = y. \]

{\scriptsize\emph{Note.} If~$A$ is reduced, then~$\nabla$ coincides
with~$\neg\neg$ and the claim reduces to statement~(\ref{item:negneg-stable})
on page~\pageref{item:negneg-stable}. In general, the operator~$\nabla$ is
related to the notion of \emph{scheme-theoretic dense open
subsets}~\cite[Lemma~9.11]{blechschmidt:phd}.\par}
\item Verify that an element of~$A^\sim$ is regular iff it
is~$\nabla$-invertible, that is:
\[ 1 \models \forall f\?A^\sim\_
  \bigl(
    (\forall s\?A^\sim\_ fs = 0 \Rightarrow s = 0) \Longleftrightarrow
    \nabla(\exists g\?A^\sim\_ fg = 1)
  \bigr)
\]
\item Verify the analogous statement for finitely generated ideals:
\begin{multline*}1 \models \forall f_1\?A^\sim\_ \ldots \forall f_n\?A^\sim\_ \\
  \bigl(
    (\forall s\?A^\sim\_ (\textstyle\bigwedge_{i=1}^n f_is = 0) \Rightarrow s = 0) \Longleftrightarrow
    \nabla(1 \in (f_1,\ldots,f_n))
  \bigr) \end{multline*}
\item Explain how the local--global principle ``if~$(f_1,\ldots,f_n)$ is a
regular ideal and if an ideal~$(g_1,\ldots,g_m)$ is regular when considered
over each localization~$A[f_i^{-1}]$, then~$(g_1,\ldots,g_m)$ is regular
over~$A$'' can be viewed as an instance of modus ponens for local operators
($(\nabla\varphi \wedge (\varphi \Rightarrow \nabla\psi)) \Longrightarrow
\nabla\psi$).
\item Fill in the details, unwind the following proof of McCoy's theorem
(page~\pageref{par:mccoy}) and compare the result with the proof in
Ref.~\cite[Theorem~2.4]{coquand-quitte:constructive-finite-free-resolutions}.

\begin{quote}
\textbf{Theorem.} Let~$M \in A^{n \times m}$ be an injective matrix.
Then~$(\Lambda^m M)$, the ideal of~$m$-minors of~$M$, is regular.

\textbf{Proof.} We verify that, under~$\nabla$, there exists an
invertible~$m$-minor, that is we verify~$\nabla(\bigvee_f (\exists g\?A^\sim\_
fg=1))$, where the disjunction is over the~$m$-minors of~$M$.

Since~$M$ is injective, the ideal generated by the first column is regular,
hence under~$\nabla$ the unit ideal. Since our goal is~$\nabla$-stable, we may assume
that it actually is the unit ideal. Hence one of the entries in the first
column is invertible. Applying elementary row and column transformations, we
can put~$M$ into the
form~$\left(\begin{smallmatrix}1&0\\0&M'\end{smallmatrix}\right)$. Continuing
in this fashion, we obtain the
form~$\left(\begin{smallmatrix}I\\0\end{smallmatrix}\right)$. This matrix
obviously has an invertible~$m$-minor, and hence the original one has one as
well.
\end{quote}
\end{alphlist}
\end{exercise}
